### 2023-09-13 doc/_variational_inference_primer.qmd
remove conditioning of data since it doesnt help

Variational inference circumvents these technical challenges by proposing an
approximate solution to the problem. Instead of working with the posterior
distribution in its full glory $\pi(\underline{\theta} \mid \underline{x})$, let
us propose an approximate posterior distribution $q_\phi$ that belongs to a
distribution family fully parametrized by $\phi$. For example, let us say that
the distribution $q_\phi$ belongs to the family of multivariate Normal
distributions such that $\phi = (\underline{\mu},
\underline{\underline{\Sigma}})$, where $\underline{\mu}$ is the vector of means
and $\underline{\underline{\Sigma}}$ is the covariance matrix. If we replace
$\pi$ by $q_\phi$, we want $q_\phi$ to resemble the original posterior as much
as possible. Mathematically, this can be expressed as minimizing a "*distance
metric*"---the Kullback-Leibler (KL) divergence, for example---between the
distributions. Note that we use quotation marks because, formally, the KL
divergence is not a distance metric since it is not symmetric. Nevertheless, the
variational objective is set to find a distribution $q_\phi^*$ such that
$$
q_\phi^*(\underline{\theta} \mid \underline{x}) =
\min_\phi D_{KL}\left(
    q_\phi(\underline{\theta} \mid \underline{x}) \vert\vert 
    \pi(\underline{\theta} \mid \underline{x})
\right),
$${#eq-vi_objective}
where $D_{KL}$ is the KL divergence. We alert the reader that in the literature
$q_\phi$ is usually  written not conditioned on the observations, i.e.,
$q_\phi(\underline{\theta})$. But we think that writing the conditioning
reflects that the selection of the distribution $q_\phi^*$ out of the parametric
family depends on having the observations $\underline{x}$ in hand. Furthermore,
we highlight that the KL divergence is a strictly positive number, i.e.,
$$
D_{KL}\left(
    q_\phi(\underline{\theta} \mid \underline{x}) \vert\vert 
    \pi(\underline{\theta} \mid \underline{x})
\right) \geq 0,
$${#eq-vi_kl_pos}
as this property will become important later on.

At first sight, @eq-vi_objective does not improve the situation but only
introduces further technical complications. After all, the definition of the KL
divergence
$$
D_{KL}\left(
    q_\phi(\underline{\theta} \mid \underline{x}) \vert\vert 
    \pi(\underline{\theta} \mid \underline{x})
\right) \equiv 
\int \cdots \int d\underline{\theta}^K
q_\phi(\underline{\theta} \mid \underline{x})
\ln \frac{
    q_\phi(\underline{\theta} \mid \underline{x})
}{
    \pi(\underline{\theta} \mid \underline{x})
},
$${#eq-vi_kl_div}
includes the posterior distribution $\pi(\underline{\theta} \mid \underline{x})$
we are trying to get around. However, let us manipulate @eq-vi_kl_div to beat it
to a more reasonable form. First, we can use the properties of the logarithms to
write
$$
D_{KL}\left(
    q_\phi(\underline{\theta} \mid \underline{x}) \vert\vert 
    \pi(\underline{\theta} \mid \underline{x})
\right) = 
\int d\underline{\theta}^K q_\phi(\underline{\theta} \mid \underline{x})
\ln q_\phi(\underline{\theta} \mid \underline{x}) -
\int d\underline{\theta}^K q_\phi(\underline{\theta} \mid \underline{x})
\ln \pi(\underline{\theta} \mid \underline{x}),
$${#eq-vi_step01}
where, for convenience, we write a single integration sign
($d\underline{\theta}^K$ still represents a multi-dimensional differential). For
the second term in @eq-vi_step01, we can substitute the term inside the
logarithm using @eq-vi_bayes_theorem. This results in
$$
\begin{aligned}
D_{KL}\left(
    q_\phi(\underline{\theta} \mid \underline{x}) \vert\vert 
    \pi(\underline{\theta} \mid \underline{x})
\right) &= 
\int d\underline{\theta}^K q_\phi(\underline{\theta} \mid \underline{x})
\ln q_\phi(\underline{\theta} \mid \underline{x}) \\
&- \int d\underline{\theta}^K q_\phi(\underline{\theta} \mid \underline{x})
\ln \left( 
    \frac{
        \pi(\underline{x} \mid \underline{\theta})\pi(\underline{\theta})
    }{
        \pi(\underline{x})
    }
\right).
\end{aligned}
$${#eq-vi_step02}
Again, using the properties of logarithms, we can split @eq-vi_step02, obtaining
$$
\begin{aligned}
D_{KL}\left(
    q_\phi(\underline{\theta} \mid \underline{x}) \vert\vert 
    \pi(\underline{\theta} \mid \underline{x})
\right) &= 
\int d\underline{\theta}^K q_\phi(\underline{\theta} \mid \underline{x})
\ln q_\phi(\underline{\theta} \mid \underline{x}) \\
&-\int d\underline{\theta}^K q_\phi(\underline{\theta} \mid \underline{x})
\ln \pi(\underline{x} \mid \underline{\theta}) \\
&-\int d\underline{\theta}^K q_\phi(\underline{\theta} \mid \underline{x})
\ln \pi(\underline{\theta}) \\
&+\int d\underline{\theta}^K q_\phi(\underline{\theta} \mid \underline{x})
\ln \pi(\underline{x}).
\end{aligned}
$${#eq-vi_step03}
It is convenient to write @eq-vi_step03 as
$$
\begin{aligned}
D_{KL}\left(
    q_\phi(\underline{\theta} \mid \underline{x}) \vert\vert 
    \pi(\underline{\theta} \mid \underline{x})
\right) &= 
\int d\underline{\theta}^K q_\phi(\underline{\theta} \mid \underline{x})
\ln \frac{
    q_\phi(\underline{\theta} \mid \underline{x})
    }{
        \pi(\underline{\theta})
    } \\
&-\int d\underline{\theta}^K q_\phi(\underline{\theta} \mid \underline{x})
\ln \pi(\underline{x} \mid \underline{\theta}) \\
&+ \ln \pi(\underline{x}) 
\int d\underline{\theta}^K q_\phi(\underline{\theta} \mid \underline{x}),
\end{aligned}
$${#eq-vi_step04}
where for the last term, we can take $\ln \pi(\underline{x})$ out of the
integral since it does not depend on $\underline{\theta}$. Lastly, we utilize
two properties:

1. The proposed approximate distribution must be normalized, i.e.,
$$
\int d\underline{\theta}^K q_\phi(\underline{\theta} \mid \underline{x}) = 1.
$${#eq-vi_q_norm}
2. The law of the unconscious statistician (LOTUS) establishes that for any
probability density function, it must be true that
$$
\int d\underline{\theta}^K q_\phi(\underline{\theta} \mid \underline{x})
f(\underline{\theta}) = \left\langle 
    f(\underline{\theta}) 
\right\rangle_{q_\phi},
$${#eq-vi_lotus}
where $\left\langle\cdot\right\rangle_{q_\phi}$ is the expected value over the
$q_\phi$ distribution.

Using these two properties, the positivity constraint on the KL divergence in
@eq-vi_kl_pos, and the definition of the KL divergence in @eq-vi_kl_div we can
rewrite @eq-vi_step04 as
$$
D_{KL}\left( 
    q_\phi(\underline{\theta} \mid \underline{x}) \vert \vert
    \pi(\underline{\theta}) 
\right) -
\left\langle
    \ln \pi(\underline{x} \mid \underline{\theta})
\right\rangle_{q_\phi}
\geq - \ln \pi(\underline{x}).
$${#eq-vi_step05}
Multiplying by a minus one, we have the functional form of the so-called
evidence lower bound (ELBO) @kingma2014,
$$
\underbrace{
    \ln \pi(\underline{x})
}_{\text{log evidence}} \geq
\underbrace{
    \left\langle
        \ln \pi(\underline{x} \mid \underline{\theta})
    \right\rangle_{q_\phi} -
    D_{KL}\left( 
        q_\phi(\underline{\theta} \mid \underline{x}) \vert \vert
        \pi(\underline{\theta}) 
    \right)
}_{\text{ELBO}}.
$${#eq-vi_elbo}

Let us recapitulate where we are. We started by presenting the challenge of
working with Bayes' theorem, as it requires a high-dimensional integral of the
form in @eq-vi_evidence_int. As an alternative, variational inference posits to
approximate the posterior distribution $\pi(\underline{\theta} \mid
\underline{x})$ with a parametric distribution $q_\phi(\underline{\theta} \mid
\underline{x})$. By minimizing the KL divergence between these distributions, we
arrive at the result in @eq-vi_elbo, where the left-hand side---the log
marginalized likelihood or log evidence---we cannot compute for
technical/computational reasons. However, the right-hand side is composed of
things we can easily evaluate. We can easily evaluate the log-likelihood $\ln
\pi(\underline{x} \mid \underline{\theta})$ and the KL divergence between our
proposed approximate distribution $q_\phi(\underline{\theta} \mid
\underline{x})$ and the prior distribution $\pi(\underline{\theta})$. Moreover,
we can compute the gradients of these functions with respect to the parameters
of our proposed distribution. This last point implies that we can change the
parameters of the proposed distribution to maximize the ELBO. And, although we
cannot compute the left-hand side of @eq-vi_elbo, we know that however large we
make the ELBO, it will always be smaller than (or equal) the log-marginal
likelihood. Therefore, the larger we can make the ELBO by modifying the
parameters $\phi$, the closer it gets to the log-marginal likelihood, and, as a
consequence, the better our proposed distribution $q_\phi(\underline{\theta}
\mid \underline{x})$ gets to the true posterior distribution
$\pi(\underline{\theta} \mid \underline{x})$.

>>>>

Variational inference circumvents these technical challenges by proposing an
approximate solution to the problem. Instead of working with the posterior
distribution in its full glory $\pi(\underline{\theta} \mid \underline{x})$, let
us propose an approximate posterior distribution $q_\phi$ that belongs to a
distribution family fully parametrized by $\phi$. For example, let us say that
the distribution $q_\phi$ belongs to the family of multivariate Normal
distributions such that $\phi = (\underline{\mu},
\underline{\underline{\Sigma}})$, where $\underline{\mu}$ is the vector of means
and $\underline{\underline{\Sigma}}$ is the covariance matrix. If we replace
$\pi$ by $q_\phi$, we want $q_\phi$ to resemble the original posterior as much
as possible. Mathematically, this can be expressed as minimizing a "*distance
metric*"---the Kullback-Leibler (KL) divergence, for example---between the
distributions. Note that we use quotation marks because, formally, the KL
divergence is not a distance metric since it is not symmetric. Nevertheless, the
variational objective is set to find a distribution $q_\phi^*$ such that
$$
q_\phi^*(\underline{\theta}) =
\min_\phi D_{KL}\left(
    q_\phi(\underline{\theta}) \vert\vert 
    \pi(\underline{\theta} \mid \underline{x})
\right),
$${#eq-vi_objective}
where $D_{KL}$ is the KL divergence. Furthermore, we highlight that the KL
divergence is a strictly positive number, i.e.,
$$
D_{KL}\left(
    q_\phi(\underline{\theta}) \vert\vert 
    \pi(\underline{\theta} \mid \underline{x})
\right) \geq 0,
$${#eq-vi_kl_pos}
as this property will become important later on.

At first sight, @eq-vi_objective does not improve the situation but only
introduces further technical complications. After all, the definition of the KL
divergence
$$
D_{KL}\left(
    q_\phi(\underline{\theta}) \vert\vert 
    \pi(\underline{\theta} \mid \underline{x})
\right) \equiv 
\int \cdots \int d\underline{\theta}^K
q_\phi(\underline{\theta})
\ln \frac{
    q_\phi(\underline{\theta})
}{
    \pi(\underline{\theta} \mid \underline{x})
},
$${#eq-vi_kl_div}
includes the posterior distribution $\pi(\underline{\theta} \mid \underline{x})$
we are trying to get around. However, let us manipulate @eq-vi_kl_div to beat it
to a more reasonable form. First, we can use the properties of the logarithms to
write
$$
D_{KL}\left(
    q_\phi(\underline{\theta}) \vert\vert 
    \pi(\underline{\theta} \mid \underline{x})
\right) = 
\int d\underline{\theta}^K q_\phi(\underline{\theta})
\ln q_\phi(\underline{\theta}) -
\int d\underline{\theta}^K q_\phi(\underline{\theta})
\ln \pi(\underline{\theta} \mid \underline{x}),
$${#eq-vi_step01}
where, for convenience, we write a single integration sign
($d\underline{\theta}^K$ still represents a multi-dimensional differential). For
the second term in @eq-vi_step01, we can substitute the term inside the
logarithm using @eq-vi_bayes_theorem. This results in
$$
\begin{aligned}
D_{KL}\left(
    q_\phi(\underline{\theta}) \vert\vert 
    \pi(\underline{\theta} \mid \underline{x})
\right) &= 
\int d\underline{\theta}^K q_\phi(\underline{\theta})
\ln q_\phi(\underline{\theta}) \\
&- \int d\underline{\theta}^K q_\phi(\underline{\theta})
\ln \left( 
    \frac{
        \pi(\underline{x} \mid \underline{\theta})\pi(\underline{\theta})
    }{
        \pi(\underline{x})
    }
\right).
\end{aligned}
$${#eq-vi_step02}
Again, using the properties of logarithms, we can split @eq-vi_step02, obtaining
$$
\begin{aligned}
D_{KL}\left(
    q_\phi(\underline{\theta}) \vert\vert 
    \pi(\underline{\theta} \mid \underline{x})
\right) &= 
\int d\underline{\theta}^K q_\phi(\underline{\theta})
\ln q_\phi(\underline{\theta}) \\
&-\int d\underline{\theta}^K q_\phi(\underline{\theta})
\ln \pi(\underline{x} \mid \underline{\theta}) \\
&-\int d\underline{\theta}^K q_\phi(\underline{\theta})
\ln \pi(\underline{\theta}) \\
&+\int d\underline{\theta}^K q_\phi(\underline{\theta})
\ln \pi(\underline{x}).
\end{aligned}
$${#eq-vi_step03}
It is convenient to write @eq-vi_step03 as
$$
\begin{aligned}
D_{KL}\left(
    q_\phi(\underline{\theta}) \vert\vert 
    \pi(\underline{\theta} \mid \underline{x})
\right) &= 
\int d\underline{\theta}^K q_\phi(\underline{\theta})
\ln \frac{
    q_\phi(\underline{\theta})
    }{
        \pi(\underline{\theta})
    } \\
&-\int d\underline{\theta}^K q_\phi(\underline{\theta})
\ln \pi(\underline{x} \mid \underline{\theta}) \\
&+ \ln \pi(\underline{x}) 
\int d\underline{\theta}^K q_\phi(\underline{\theta}),
\end{aligned}
$${#eq-vi_step04}
where for the last term, we can take $\ln \pi(\underline{x})$ out of the
integral since it does not depend on $\underline{\theta}$. Lastly, we utilize
two properties:

1. The proposed approximate distribution must be normalized, i.e.,
$$
\int d\underline{\theta}^K q_\phi(\underline{\theta}) = 1.
$${#eq-vi_q_norm}
2. The law of the unconscious statistician (LOTUS) establishes that for any
probability density function, it must be true that
$$
\int d\underline{\theta}^K q_\phi(\underline{\theta})
f(\underline{\theta}) = \left\langle 
    f(\underline{\theta}) 
\right\rangle_{q_\phi},
$${#eq-vi_lotus}
where $\left\langle\cdot\right\rangle_{q_\phi}$ is the expected value over the
$q_\phi$ distribution.

Using these two properties, the positivity constraint on the KL divergence in
@eq-vi_kl_pos, and the definition of the KL divergence in @eq-vi_kl_div we can
rewrite @eq-vi_step04 as
$$
D_{KL}\left( 
    q_\phi(\underline{\theta}) \vert \vert
    \pi(\underline{\theta}) 
\right) -
\left\langle
    \ln \pi(\underline{x} \mid \underline{\theta})
\right\rangle_{q_\phi}
\geq - \ln \pi(\underline{x}).
$${#eq-vi_step05}
Multiplying by a minus one, we have the functional form of the so-called
evidence lower bound (ELBO) @kingma2014,
$$
\underbrace{
    \ln \pi(\underline{x})
}_{\text{log evidence}} \geq
\underbrace{
    \left\langle
        \ln \pi(\underline{x} \mid \underline{\theta})
    \right\rangle_{q_\phi} -
    D_{KL}\left( 
        q_\phi(\underline{\theta}) \vert \vert
        \pi(\underline{\theta}) 
    \right)
}_{\text{ELBO}}.
$${#eq-vi_elbo}

Let us recapitulate where we are. We started by presenting the challenge of
working with Bayes' theorem, as it requires a high-dimensional integral of the
form in @eq-vi_evidence_int. As an alternative, variational inference posits to
approximate the posterior distribution $\pi(\underline{\theta} \mid
\underline{x})$ with a parametric distribution $q_\phi(\underline{\theta})$. By
minimizing the KL divergence between these distributions, we arrive at the
result in @eq-vi_elbo, where the left-hand side---the log marginalized
likelihood or log evidence---we cannot compute for technical/computational
reasons. However, the right-hand side is composed of things we can easily
evaluate. We can easily evaluate the log-likelihood $\ln \pi(\underline{x} \mid
\underline{\theta})$ and the KL divergence between our proposed approximate
distribution $q_\phi(\underline{\theta})$ and the prior distribution
$\pi(\underline{\theta})$. Moreover, we can compute the gradients of these
functions with respect to the parameters of our proposed distribution. This last
point implies that we can change the parameters of the proposed distribution to
maximize the ELBO. And, although we cannot compute the left-hand side of
@eq-vi_elbo, we know that however large we make the ELBO, it will always be
smaller than (or equal) the log-marginal likelihood. Therefore, the larger we
can make the ELBO by modifying the parameters $\phi$, the closer it gets to the
log-marginal likelihood, and, as a consequence, the better our proposed
distribution $q_\phi(\underline{\theta})$ gets to the true posterior
distribution $\pi(\underline{\theta} \mid \underline{x})$.
